\begin{abstract}
The European Statistical System (ESS) is increasingly integrating non-traditional data sources and machine learning methods to enhance statistical production. These sources promise greater timeliness and spatiotemporal resolution but bring complexities and challenges, particularly in coding, classification, data editing, and imputation, addressed in the Bucharest Memorandum. Innovations in machine learning are at the forefront, yet the integration into National Statistical Offices (NSOs) is hampered by inadequate IT infrastructures and a lack of necessary skills within traditional statistical environments. This paper discusses the development of Onyxia, an open-source project aimed at enabling the deployment of data science platforms that enhance the autonomy of statisticians and embody the Trusted Smart Statistics initiative. We explore the technological shifts within the data ecosystem that led to the adoption of cloud-native technologies, such as containers and object storage, and how these technologies contribute to building modern, scalable, and flexible statistical environments. We contend that the autonomy to innovate is crucial for statisticians in the data science era and that overcoming IT infrastructure and skill constraints is key to fostering this independence. In detail, we address how Onyxia facilitates access to cloud resources for statisticians through a user-friendly interface, circumventing the complexity of cloud technology configuration and potential vendor lock-in. The establishment of SSP Cloud, an open-innovation instance of Onyxia, showcases how these platforms can be adopted and the new development practices they promote. Finally, we present a case study of the classification of French companies' activities, demonstrating how Onyxia’s infrastructure and principles were vital in operationalizing a machine learning model at Insee in accordance with MLOps principles, marking a milestone in the institution's innovation strategy. 
\end{abstract}

\keywords{Official Statistics, Big Data, Cloud Technologies, Data Science, Machine Learning, MLOps, Reproducibility, Open Source}
