\section{Discussion}

% - Global evolution

Initially developed as an internal project, Onyxia has gained recognition beyond the realm of Insee or the French administration. Aware of the necessity to foster autonomy in order to leverage the full potential of data science, several organizations now have a production instance of Onyxia running, and multiple others are in the process of either testing or implementing one. Besides, the choice of Onyxia as the reference data science platform in the context of the One-Stop-Shop for Artificial Intelligence/Machine Learning for Official Statistics (AIML4OS) will further facilitate its adoption within the ESS. This trend is naturally very beneficial to the Onyxia project, as it moves from a project developed in open-source - but mainly at Insee - to a full open-souce project with a growing base of contributors. This in turn facilitates its adoption by other organizations, since it gives more guarantees on its sustainability independently from Insee's strategy. The governance of the project is currently evolving to reflect this trend. For instance with the organization of monthly community calls and the creation of a public channel and roadmap for the project\footnote{All information are available on the GitHub depository of the project : \url{https://github.com/InseeFrLab/onyxia}}.

Despite this success, we observe several limitations to the widespread adoption of the project in organizations. First, it is essential to remind that the fundamental choice made by organizations that adopt Onyxia is not the software itself, but the underlying technologies: containerization (through Kubernetes) and object storage. These technologies can represent substantial entry costs for organizations, as they demand a significant commitment to developing and maintaining skills which are not readily found in NSOs. Yet, the general trend towards cloud-native solutions among data-centric organizations suggests a favorable shift that could mitigate these challenges over time.

Similarly, the transition towards cloud-native technologies induces entry costs for statisticians. First, they often deal with a loss of references regarding where computations actually happen: while they may be accustomed to performing computation on centralized servers rather than a personal computer, the container adds a layer of abstraction that make the location hard to grasp at first. But the major perceived change in this paradigm is the loss of data persistence. In traditional setups - either a personal computer or a server accessed through a virtual desktop - the code, the data and the computing environment are kind of mixed in a black box fashion. On the contrary, containers have no persistence by design. While object storage provides this persistence, a proper use of these infrastructures for statistical projects require a variety of tools and corresponding skills: using a version control system for the code, interacting with the object storage API to store the data, providing configuration files or secrets as inputs, etc. In a way, these entry costs can be seen as the "price" of autonomy: thanks to cloud-native technologies, statisticians now have access to scalable and flexible environments that enable them to experiment more freely, but this autonomy requires a significant skills upgrade which may be overwhelming at first and limit adoption. However, our experience at Insee suggests that this effect can largely be mitigated through a combination of training statisticians to development best practices and accompanying statistical projects when transitioning to cloud infrastructures.

While Onyxia has significantly democratized access to cloud-native technologies for statisticians, the journey towards integrating data science within NSOs encompasses broader organizational challenges beyond the technical realm. In particular, the deployment of our first machine learning model in production highlighted the necessity of overcoming skill compartmentalization across IT, business, and innovation teams. Effective collaboration and the seamless integration of machine learning into statistical processes require not only shared goals and methodologies but also a convergence of diverse skills and knowledge. Addressing these challenges involves strategic measures such as embedding data science capabilities within business teams for better alignment with project objectives, and initiating cross-disciplinary training to harmonize the toolsets and languages across teams. Ultimately, the successful transition to a data science-driven approach in statistical production is contingent upon a balanced strategy that marries technical solutions like Onyxia with comprehensive organizational adjustments, fostering a culture of collaboration, continuous learning, and innovation.

% \subsubsection{Governance and collaboration challenges in ML Projects}

% During the deployment of our first model into production, we encountered several governance challenges that we had not anticipated. As explained in Section \ref{sec:mlops}, we initiated the project with three distinct teams: an IT team comprised of developers, the core business team responsible for maintaining the Sirene directory, and finally, the innovation team consisting of data scientists and data engineers. The creation of the SSP Lab\footnote{Name of the innovation team.} at Insee in 2018 was justified to support the business teams on projects with relatively precise objectives in terms of statistical production and to experimentally introduce new methods into the statistical process, including machine learning methods. Initially, the innovation team's sole objective was to conduct an experiment and let the business team judge its relevance based on the results obtained. However, due to the urgency and quality of the results, the innovation team continued to collaborate with the business team to assist in production deployment.

% This phase highlighted several issues, with the primary one being the compartmentalization of skills. The innovation team had limited knowledge of business issues and was unable to make certain decisions independently. The business team, on the other hand, had very few data science skills and was unable to manage the production deployment of the experiment alone. Finally, the IT team, although well-versed in DevOps best practices, had no knowledge of MLOps approach and of the methodological tools used in machine learning. Additionally, the programming languages used by the production and innovation teams were different, which have slowed down the model's production deployment. Before using MLflow and its Model Registry functionality, the preprocessing done in Python had to be reprogrammed in Java, which proved to be very tedious. We realized how code duplication was a source of errors and needed to be avoided as much as possible.

% To address these challenges, several measures were implemented at Insee. Firstly, a data scientist was hired in September 2023 and integrated into the business team to take responsibility for the model in production, its monitoring, and retraining. The goal is to have someone who fully understands the business issues to quickly integrate new features as needed. Furthermore, to anticipate the arrival of multiple machine learning models in production, a Python training plan for developers was launched to enhance their skills with open-source software and align development and production languages.
